Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-02-05 paper/draft.tex

\documentclass[preprint,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded). Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image with a 200nm diameter quantum dot coupled to the right side. The channel can be heated by driving current through the quantum point contact on the left. Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b) Calculated charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from the shift in $V_0$ which relates to a change in entropy between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg$ is anti-symmetric about $V_0$.}
\end{figure}

@@

In order to accomplish this entropy to charge conversion, we make use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, $S$, (as well as particle number, $N$, and temperature, $T$) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition of Eq.~\ref{eqn:max} is effectively met by working well below the Fermi temperature, $T_F \sim$\SI{5}{\kelvin}, where the pressure is simply the degeneracy pressure \cite{Landau1969}. Refs.~\onlinecite{Cooper2009} and ~\onlinecite{Ben-Shach2013} propose schemes to measure the entropy derivative, $\frac{\partial S}{\partial N}$, of the $\nu = \frac{5}{2}$ state via this Maxwell relation, detecting changes in the quasiparticle chemical potential with temperature.  Our work builds on these ideas to measure the entropy of the few-electron ground states of a quantum dot. Unlike the $\nu = \frac{5}{2}$ quasiparticles described in the theoretical work, the few-electron quantum dot states studied here have an entropy that is well-understood from the spin degree of freedom \cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}. Our experiment demonstrates a direct entropy measurement on a few-particle system; in the future, the method can be easily extended as a probe into the nature of more exotic systems, such as topologically non-trivial Majorana and other non-Abelian states.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a realizes an electron reservoir (central region) in thermal equilibrium with a few-electron quantum dot coupled to its right side, whose occupation is measured using an adjacent charge sensing quantum point contact\cite{Staring2007, Frolov2009, Thierschmann2015}. The number of electrons on the dot can be tuned using the gate voltage, $V_G$, to control the chemical potential.  An electron is added to an $N$ electron dot when the chemical potential to add an electron, $\mu_{N+1}$, drops below the Fermi level of the reservoir, $E_F$.  Figure 1b illustrates a such a transition, showing the expected step in charge sensor conductance as a function of $V_G$ and thermal broadening by the reservoir temperature, $T_{res}$. The gate voltage corresponding to midpoint of the transition, $V_0$, marks the chemical potential where the probabilities of finding $N$ or $N+1$ electrons on the dot are equal.

Any entropy change from the $N$ to $N+1$ ground states of the dot should shift the midpoint of the transition, $V_0$, by changing the chemical potential on the dot as in Eq.~\ref{eqn:max}.  Detecting the shift in $V_0$ with respect to temperature (${\propto}\frac{\partial \mu}{\partial T}$), amounts to a measurement of the entropy difference $dS$ between the $N$ to $N+1$ ground states ($dN=1$).  In practice, the shift in $V_0$ with temperature is measured by oscillating $T_{res}$ and monitoring any resulting changes in $g_{sens}$ with a lockin amplifier.  If the transition is broadened but not shifted due to heating ($dS=0$), an antisymmetric lineshape is expected from this lockin measurement.  If the transition is not only broadened but also shifted ($dS,d\mu\neq0$), the change in $g_{sens}$ due to heating will asymmetric, as shown in Fig.~\ref{fig:fig1}c, thus offering a clearly distinguishable signature of entropy for this single charge measurement.

An explanation based on detailed balance is helpful in understanding how the shift in $\mu_{N+1}$ (with respect to $E_F$) upon heating relates to entropy.   In general, electrons tunnel back and forth between the dot and reservoir as long as $\mu_{N+1}$ is within thermal broadening of Fermi level, that is, as long as there are available states in both the dot and the reservoir.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N\rightarrow N+1}$ and $\Gamma_{out}=\Gamma_{N+1\rightarrow N}$, depend on degeneracy of the $N$ and $N+1$ ground states.  Consider the filling of the first electron state at zero magnetic field, a state with a single, unpaired spin-$\frac{1}{2}$ and therefore a degeneracy of 2. An electron tunneling into this state has both spin-up and -down states available.  Tunneling out, the electron (whether spin up or down) must tunnel into reservoir states with the same spin, that is, only half of the reservoir states are available. As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N+1}=E_F$. Extending this logic to a general case gives the same results as derived from detailed balance \cite{Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N+1} \Gamma f(E_F - \mu_{N+1}) \label{eqn:rates}\\
	\Gamma_{out} &= g_{N} \Gamma [1 - f(E_F - \mu_{N+1})] \nonumber
\end{align}
%
where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N/N+1}$ are the degeneracies of the $N$ and $N+1$ electron states.

The midpoint of the transition between the $N$ and $N+1$ electron states, measured by the charge sensor, occurs when the dot is occupied half the time, that is, when $\Gamma_{in} = \Gamma_{out}$. From Eq.~\ref{eqn:rates}, $\Gamma_{in} = \Gamma_{out}$ occurs when $\mu_{N+1} = E_F$ as long as $g_{N}=g_{N+1}$. When $g_{N}\neq g_{N+1}$, the midpoint of the transition shifts away from $\mu_{N+1}=E_F$ by an amount that depends on the broadening of the Fermi function, and therefore on $T$.   Thus, a measurement of how the midpoint of the transition ($V_0$) shifts with T allows the ratio $g_{N+1}/g_{N}$ to be determined, which itself is related to $dS$ through the Boltzmann entropy formula, $S_{N}=k_{B} \ln{g_N}$.  The relationship between degeneracies and tunnel rates was recently explored in a time-resolved measurement \cite{Hofmann2016}.  The thermal technique presented here extends the measurement of entropy (degeneracy) to a wider set of applications, where the tunneling process may not be convenient or even possible to observe directly.

%%% fab/measurement details %%%
The device was built on a AlGaAs/GaAs heterostructure, hosting a 2D electron gas with density and mobility at \SI{300}{\milli\kelvin} (determined on a separate chip) of \SI{2.42e11}{\per\square\centi\metre} and \SI[per-mode=symbol]{2.56e6}{\square\centi\metre\per\volt\per\second}.   Mesas and NiAuGe ohmic contacts to the 2DEG were defined by standard photolithography techniques, then \SI{10}{\nano\metre} of $\mathrm{HfO_2}$ was deposited via atomic layer deposition to improve the gating stability to the device. Electron beam lithography followed by evaporation of \SI{20}{\nano\metre} of Ti/Au defined the fine gate structures, including the \SI{200}{\nano\metre} diameter quantum dots. The measurement was carried out in an dilution refrigerator with a base temperature of \SI{14}{\milli\kelvin}.

@@

%%% device setup %%%
Using the gates on the right side of Fig~\ref{fig:fig1}a, the quantum dot was tuned such that the source was tunnel coupled to the reservoir with the drain closed. The charge sensing QPC was tuned to a conductance of ${\sim}e^2/h$, where conductance was most sensitive to charge on the dot. A small DC voltage bias was applied across the QPC and the measured current was used to calculate the sensor conductance. The reservoir temperature was controlled by varying the mixing chamber temperature or Joule heating through the QPC on the left side of the channel. In the latter case, the QPC was fixed at \SI{20}{\kilo\ohm} and biased with an AC or DC current source. The Joule heating setup gives fast control of the temperature (up to \SI{100}{\hertz}) while keeping the voltage drop across the reservoir negligibly small.

%%% figs 2a and b %%%
Fig.~\ref{fig:fig2}a shows the charge sensor conductance as the first electron is added to the dot by tuning $V_G$, at two different temperatures. The higher temperature transition curve is noticeably broadened by the increased temperature. Assuming the source barrier is sufficiently large, $\Gamma_{in/out}$ are broadened only by the temperature of the reservoir and the conductance takes the form,
%
\begin{align}
\label{eqn:g}
        g_{sens}(V,\Theta) &= G_0 \tanh\left(\frac{V-V_0(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1\left[V-V_0(\Theta)\right] + G_2 \nonumber
\end{align}
%
where $\Theta = \frac{k_B T}{\alpha e}$, $\alpha$ is the lever arm for the quantum dot plunger gate, and $G_1$ is determined by the cross capacitance between the sensor gates and the plunger gate. The dependence of $V_0$ on $\Theta$ captures any shift in $\mu$ due to a change in entropy (degeneracy) across the transition. Fits to Eq.~\ref{eqn:g} are seen, in addition to the sensor data, in Fig.~\ref{fig:fig2}a. From these fits the transition width, $\Theta$, can be extracted. In Fig.~\ref{fig:fig2}b $\Theta$ is plotted as a function of mixing chamber temperature, $T_{MC}$. From this plot, it is clear that Eq.~\ref{eqn:g} holds when $T_{MC}\gtrsim$ \SI{100}{\milli\kelvin}. For this reason, the mixing chamber temperature remains fixed at \SI{100}{\milli\kelvin} for the remainder of this work. Fitting high temperature data in Fig.~\ref{fig:fig2} to the definition of $\Theta$ (with $T=T_{MC}$) yields $\alpha$, making it possible to convert from $V_G$ to $\mu$.

%%% figs 2c and d%%%
Data in Fig.~\ref{fig:fig2}c, and the corresponding fits, show how $dS$ was measured across a single charge transition. Fluctuations in $g_{sens}$, were measured as $T_{res}$ was varied using Joule heating with an AC current source ($f_{Heat} =$ \SI{48.7}{\hertz}). The change in $T_{res}$ is given by $dT \sim I^{AC}_{Heat} \sim \sin^2(2\omega t)$. A calibration for $dT$, using $\alpha$ as measured in Fig.~\ref{fig:fig2}b, can be seen in Fig.~\ref{fig:fig2}c. To avoid thermally accessing excited states of the dot it must be true that $k_B dT \ll \Delta$ where $\Delta$ is the level spacing between the ground and first excited states; which holds for all transitions investigated.

To acquire $dg_{sens}$, current fluctuations through the charge sensor were measured at $2f_{Heat}$ using a lockin amplifier. Taking a derivative of Eq. \ref{eqn:g} with respect to temperature gives the form of $dg_{sens}$ near a charge transition,
%
\begin{align}
\label{eqn:dg}
        dg_{sens}(V, \Theta) &\propto \left[ \frac{V-V_0(\Theta)}{2\Theta} + \delta \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V-V_0(\Theta)}{2\Theta}\right) \nonumber
\end{align}
%
where $\delta=\frac{\partial V_0}{\partial \Theta}$. Using the lever arm to convert $V_0$ to a chemical potential $\mu$, the parameter $\delta$ can be related to the change in entropy through Eq.~\ref{eqn:max},
%
\begin{align}
\label{eqn:delta}
        \delta &= \frac{\partial V_0}{\partial \Theta} = 
        \frac{1}{k_B} \frac{\partial \mu}{\partial T} = 
        -\frac{1}{k_B} \partial S
\end{align}
%
Fits to Eq.~\ref{eqn:dg} are shown for $\delta=0$ and a best fit to $\delta$. From these fits it is clear $\delta\neq0$ is necessary to properly fit the asymmetry in peak height, with the best fit being $\delta \approx \ln{2}$. For an N-electron macrostate on the quantum dot, entropy is given by the Boltzmann entropy,
%
\begin{align}
\label{eqn:S}
        S &= k_B \ln{\Omega}
\end{align}
%
where $\Omega$ is the spin degeneracy of the ground state, equal to the number of microstates available. For the 0-1 transition, $dS =  k_B\ln{2} - k_B \ln{1} = k_B\ln{2}$.

%%% fig 2e %%%
Note that $\delta$ is a fit parameter independent of $dT$ and $\alpha$. Fig.~\ref{fig:fig2}e shows that $dS$, as determined from fits to $dg_{sens}$ data, remains constant over a range of $dT$ ($I^{AC}_{Heat}$). The vertical scale on the right side of Fig. \ref{fig:fig2}e, $\delta$ in units of gate voltage per unit temperature, illustrates the necessity of the lockin measurement; device stability and small signals made it difficult for $delta$ to be determined directly from $g_{sens}$ at different $T$. Without any additional calibration, it is possible to measure $dS$ directly from the lockin amplifier signal, $dg_{sens}$.

@@

To confirm that the measured signal derives from the electron spin, the behavior of $dS$ was investigated as a function of parallel magnetic field. Fig.~\ref{fig:fig3}c shows $dS$ as a function of applied field for the 0-1 and 2-3 electron transitions. The 2-electron ground state is a spin singlet, while the 3-electron ground state is same singlet with an additional unpaired electron, making the field-dependent behavior of these transitions identical where $g \mu_{B} B$ is less than the excited state energies. Zeeman splitting, measured directly in Fig. \ref{fig:fig2}b, lifts the spin degeneracy of the unpaired electrons and the entropy of the 1- and 3- electron ground states goes to zero with increasing field

The behavior of $dS$ as a function of magnetic field can be understood using the Gibbs entropy,
%
\begin{align}
\label{eqn:gibbs}
        S &= k_B \sum_{i=\pm} p_{i}(B, T) \ln{ p_{i}(B,T) }
\end{align}
%
where $p_{i}(B, T)$ is the probability of the system being in the i$^{th}$ spin state at a given field and temperature.  For states with a single unpaired electron, $p_{i}(B, T)$ can be written as a two-level system,
%
\begin{align}
\label{eqn:p}
        p_{\pm}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B}{k_B T}}}
\end{align}
%
where $\pm$ represent the spin up/down states. Fits to Eqs.~\ref{eqn:gibbs} and \ref{eqn:p} are shown in Fig \ref{fig:fig3}c for each transition. The fits allow for peak height, g-factor, and a vertical offset as free parameters with $T$ fixed at $T_{MC}$. From these fits we find $dS(B=0)=0.92 \ln{2}$ and $g=-0.33$ for the 0-1 transition.  Discrepancy between $g$ and the result shown in Fig. \ref{fig:fig3}b, is understood from the change in shape of the quantum dot when tuning between the $g_{dot}$ and $g_{sens}$ measurements [cite?]. The entropy agrees well with the expected $k_{B} \ln{2}$ at zero-field and goes to zero as expected with $B_{\parallel}$

@@

The low field (\SI{<5}{\tesla}) data in Fig. \ref{fig:fig4}c for the 1-2 electron transition is understood as the inverse of Fig \ref{fig:fig3}c. The 2-electron ground state remains in the spin singlet state with zero entropy while 1-electron ground state entropy goes from $k_B\ln{2}$ to zero due to Zeeman splitting. At high fields, the 1-electron ground state remains non-degenerate, while the 2-electron ground state becomes two-fold degenerate at \SI{9}{\tesla} where the singlet ($S$) and triplet ($T_+$) states cross. The singlet-triplet crossing is seen clearly in Fig.~\ref{fig:fig4}b.

The entropy for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities given by Eq.~\ref{eqn:p} for the 1-electron states and the following probabilities for the 2-electron singlet-triplet states,
%
\begin{align}
\label{eqn:pst}
        p_{S/T}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B - \Delta_{ST}}{k_B T}}}
\end{align}
%
where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. Fitting the data in Fig.~\ref{fig:fig4}c we find $g(B{=}0T)=-0.46$, $g(B{=}9T)=-0.31$, $\Delta_{ST}(B{=}9T)$ = \SI{240}{\micro\electronvolt}, $dS(B{=}0T) = 1.04\ln{2}$, and $dS(B{=}9T) = 1.01\ln{2}$. Each of these values is consistent with data in Figs.~\ref{fig:fig4}a and b as well as previous GaAs dot measurements [cite why g should change in high field]. This fit confirms that the $dS$ signal can be related directly to a tunable degeneracy in the system.

@@

>>>>

\documentclass[twocolumn,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded). Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image, with a 200nm diameter quantum dot coupled to the right side. The channel can be heated by driving current through the quantum point contact on the left. Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b)  Charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from a shift in $V_0$ that would arise from a change in entropy between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg$ is anti-symmetric about $V_0$.}
\end{figure}

@@

In order to accomplish this entropy to charge conversion, we make use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, $S$, (as well as particle number, $N$, and temperature, $T$) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition of Eq.~\ref{eqn:max} is effectively met by working well below the Fermi temperature, $T_F \sim$\SI{5}{\kelvin}, where the pressure is simply the degeneracy pressure \cite{Landau1969}.

This Maxwell relation forms the basis of two exciting proposals to measure the entropy derivative, $\frac{\partial S}{\partial N}$, of the $\nu = \frac{5}{2}$ state by detecting changes in the quasiparticle chemical potential with temperature, though these have not yet been realized in experiment \cite{Cooper2009,Ben-Shach2013}.  Our work builds on these proposals to measure the entropy of the few-electron ground states of a quantum dot. Unlike the $\nu = \frac{5}{2}$ quasiparticles described in the theoretical work, the few-electron quantum dot states studied here have an entropy that is well-understood from the spin degree of freedom \cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}. Our experiment demonstrates a direct entropy measurement on a few-particle system; in the future, the method can be easily extended as a probe into the nature of more exotic systems, such as topologically non-trivial Majorana and other non-Abelian states.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a uses electrostatic gates to realize an electron reservoir (central region; hereafter ``reservoir") in thermal and electrochemical equilibrium with a few-electron quantum dot (``dot") coupled to its right side.  The occupation of the dot is measured using an adjacent quantum point contact (``charge sensor")\cite{Staring2007, Frolov2009, Thierschmann2015}.  The $N+1^{\rm st}$ electron is added to an $N$ electron dot when the chemical potential, $\mu_{N+1}$, tuned by the gate voltage, $V_G$, drops below the Fermi level of the reservoir, $E_F$.

Figure 1b illustrates a such a transition---a step in the charge sensor conductance as a function of $V_G$---thermally broadened by the reservoir temperature, $T_{res}$.  The gate voltage corresponding to midpoint of the transition, $V_0$, marks the chemical potential where the probabilities of finding $N$ or $N+1$ electrons on the dot are equal.
%In practice, the shift in $V_0$ with temperature is measured by oscillating $T_{res}$ and monitoring any resulting changes in $g_{sens}$ with a lockin amplifier.  If the transition is broadened but not shifted due to heating ($dS=0$), an antisymmetric lineshape is expected from this lockin measurement.  If the transition is not only broadened but also shifted ($dS,d\mu\neq0$), the change in $g_{sens}$ due to heating will asymmetric, as shown in Fig.~\ref{fig:fig1}c, thus offering a clearly distinguishable signature of entropy for this single charge measurement.
According to Eq.~\ref{eqn:max}, the chemical potential and therefore $V_0$ shifts with temperature by an amount that is proportional to the entropy difference between the $N$ to $N+1$ ground states ($\delta S_{N\rightarrow N+1}$ for $\delta N=1$).  This shift is the quantity probed directly in the experiment described below; from these measurements the entropy of the first three energy levels in the quantum dot are built up, one by one.  Before turning to the experiment in detail, it is helpful to consider the same physics from an alternate point of view.

The relation between entropy and a temperature-induced shift in $\mu_{N+1}$ (with respect to $E_F$) can also be understood in the language of detailed balance.   Generically, the $N+1^{\rm st}$ electron tunnels back and forth between an $N$-electron dot and a reservoir as long as there are available states in both dot and reservoir, that is, as long as $\mu_{N+1}$ is within thermal broadening of $E_F$.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N\rightarrow N+1}$ and $\Gamma_{out}=\Gamma_{N+1\rightarrow N}$, depend on degeneracy of the $N$ and $N+1$ ground states.  Consider the filling of the first electron state at zero magnetic field, a state with a single, unpaired spin-$\frac{1}{2}$ and therefore a degeneracy of 2. An electron tunneling in has both spin-up and -down states available.  Tunneling out, the electron (whether spin-up or -down) must tunnel into reservoir states with the same spin, that is, only half of the reservoir states are available. As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N+1}=E_F$. Extending this logic to a general case gives the result derived from detailed balance \cite{Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N+1} \Gamma f(E_F - \mu_{N+1}) \label{eqn:rates}\\
	\Gamma_{out} &= g_{N} \Gamma [1 - f(E_F - \mu_{N+1})] \nonumber
\end{align}
%
where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N(N+1)}$ is the degeneracy of the $N(N+1)$ electron state.

The midpoint of the transition between the $N$ and $N+1$ electron states ($V_0$), as measured by the charge sensor, occurs when the dot is occupied half the time, that is, when $\Gamma_{in} = \Gamma_{out}$. From Eq.~\ref{eqn:rates}, $\Gamma_{in} = \Gamma_{out}$ corresponds to $\mu_{N+1} = E_F$ when $g_{N}=g_{N+1}$, but shifts away when $g_{N}\neq g_{N+1}$ by an amount that depends on the broadening of the Fermi function, and therefore on $T$.   A measurement of how $V_0$ shifts with T allows the ratio $g_{N+1}/g_{N}$ to be determined, which itself is related to $dS_{N\rightarrow N+1}$ through the Boltzmann entropy formula, $S_{N}=k_{B} \ln{g_N}$.  The relationship between degeneracies and tunnel rates was recently explored in a time-resolved measurement \cite{Hofmann2016}.  The thermal technique presented here extends the measurement of entropy (degeneracy) to a wider set of applications, where the tunneling process may not be convenient or even possible to observe in real-time.

%%% fab/measurement details %%%

@@

%%% device setup %%%
We now turn to the details of the experiment.  The quantum dot (Fig.~1a) was tuned such that the source was weakly tunnel coupled to the reservoir with the drain closed. The conductance of the charge sensor,  $g_{sens}$, was measured in a DC voltage biased configuration, and tuned to $g_{sens}{\sim}e^2/h$ where it was maximally sensitive to charge on the dot.  The reservoir temperature could be rapidly controlled (up to at least  \SI{100}{\hertz}) by by Joule heating through the QPC on the left side (Fig.~1a), which was fixed at \SI{20}{\kilo\ohm} and biased with an AC or DC current source.

%%% figs 2a and b %%%
The addition of the first electron to the dot was marked by a decrease in $g_{sens}$ that is well fit by the expected form for a thermally-broadened transition (Fig.~\ref{fig:fig2}a):
%by tuning $V_G$, at two different temperatures. The higher temperature transition curve is noticeably broadened by the increased temperature. Assuming the source barrier is sufficiently large, $\Gamma_{in/out}$ are broadened only by the temperature of the reservoir and the conductance takes the form,

%
\begin{align}
\label{eqn:g}
        g_{sens}(V,\Theta) &= G_0 \tanh\left(\frac{V-V_0(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1\left[V-V_0(\Theta)\right] + G_2 \nonumber
\end{align}
%
where $G_0$ quantifies the sensor sensitivity, $\Theta = \frac{k_B T}{\alpha e}$ is the thermal broadening expressed in units of gate voltage, $\alpha$ is the lever arm that converts $V_G$ to quantum dot potentials, and $G_1$ is determined by the cross capacitance between the sensor and the plunger gate. The dependence of $V_0$ on $\Theta$ captures any shift in $\mu$ due to a change in entropy across the transition.
%Fits to Eq.~\ref{eqn:g} are seen, in addition to the sensor data, in Fig.~\ref{fig:fig2}a.

The transition width, $\Theta$, extracted from fits to Eq.~\ref{eqn:g} follows the mixing chamber temperature, $T_{MC}$, down to below 100 mK (Fig.~\ref{fig:fig2}b)
%; $\alpha$ can be calculated based on this linear trend in the high temperature limit.
The assumption of thermally broadened transitions greatly simplifies the data analysis, so $T_{MC}$ is fixed at \SI{100}{\milli\kelvin} for most of this work, with Joule heating giving reservoir temperatures up to \SI{160}{\milli\kelvin}, still cold enough that no orbital levels above the ground state were excited.
%Fitting high temperature data in Fig.~\ref{fig:fig2} to the definition of $\Theta$ (with $T=T_{MC}$) yields $\alpha$, making it possible to convert from $V_G$ to $\mu$.

%%% figs 2c and d%%%
The data in Fig.~\ref{fig:fig2}c, and corresponding fits, offer an example of how $dS_{0\rightarrow 1}$ was measured across the $0 \rightarrow 1$ transition. A comparison of the 100 and 160 mK data in Fig.~\ref{fig:fig2}a shows that $g_{sens}$ drops as the temperature rises before the midpoint of the transition ($V_G-V_0<0$), whereas the conductance is higher in the warmer curve for $V_G-V_0>0$.   Oscillating the reservoir temperature by $dT_{res}$ using an AC current, $I^{AC}_{Heat}$, through the heater QPC, while locking in to resultant oscillating $dg_{sens}$ yields the characteristic peak-dip structure seen in Fig.~\ref{fig:fig2}c.  The expected lineshape of such a curve is given by the derivative of Eq. \ref{eqn:g} with respect to temperature, expressed as $dg_{sens}$ due to a temperature change $dT$:

%If $V_0$ did not change with $T_{res}$, one would expect a perfectly antisymmetric lineshape.  When $V_0$ changes with  $T_{res}$, on the other hand, as expected when there is an entropy change between the two charge states, the lineshape is not perfectly antisymmetric: this is the case shown in Fig.~\ref{fig:fig2}c.

%
\begin{align}
\label{eqn:dg}
        dg_{sens}(V_G, \Theta) &\propto dT \left[ \frac{V_G-V_0}{2\Theta} +\delta \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V_G-V_0}{2\Theta}\right) \nonumber
\end{align}
%
where $\delta=\frac{\partial V_0}{\partial \Theta}$. Using the lever arm to convert $V_0$ to chemical potential, one sees that the parameter $\delta$ is simply the change in entropy scaled by Boltzmann's constant:
%
\begin{align}
\label{eqn:delta}
        \delta &= \frac{\partial V_0}{\partial \Theta} = 
        \frac{1}{k_B} \frac{\partial \mu}{\partial T} = 
        -\frac{1}{k_B} \partial S_{N\rightarrow N+1}
\end{align}
%

From Eq.~\ref{eqn:dg}, $dg_{sens}(V_G)$ would perfectly antisymmetric around $V_0$ if $\partial S_{N\rightarrow N+1}$ were zero.  Fits to the data in Fig.~\ref{fig:fig2}c illustrate clearly that $\partial S_{0\rightarrow 1}=0$ is inconsistent with the data;  the best fit is found for $\partial S_{0\rightarrow 1}=0.71\times k_B$.  Indeed, this is what would be expected for the $0 \rightarrow 1$ electron transition at zero magnetic field: $S_0$ for the empty state should be $k_B \ln{1}=0$ whereas for the spin degenerate 1-electron state one expects $S_1=k_B\ln{2}$, giving $dS_{0\rightarrow 1}\equiv S_1 = S_0 =k_B\ln{2}$.%%% fig 2e %%%
Note that $\delta$ is a fit parameter that is independent of $dT$ and $\alpha$. Fig.~\ref{fig:fig2}e shows that $dS$,  determined from fits to data analogous to Fig.~\ref{fig:fig2}c, remains constant over a range of $dT$ ($I^{AC}_{Heat}$).  Although the extracted $\delta$ can be converted to gate-voltage-per-temperature units using a value for $\alpha$ measured separately (right axis of Fig.~\ref{fig:fig2}e), the entropy itself, $dS$, is determined directly from the fit to $g_{sens}$ without any additional calibration.
%The vertical scale on the right side of Fig. \ref{fig:fig2}e converts $\delta$ to units of gate voltage per unit temperature; the tiny magnitude of this signal  the necessity of the lockin measurement; device stability and small signals made it difficult for $delta$ to be determined directly from $g_{sens}$ at different $T$.

@@

Confirmation that the measured $dS$ derives from spin degeneracy can be seen in its evolution with in-plane magnetic field, $B_\parallel$. Figure \ref{fig:fig3}a [REMOVE PANELS AB] compares $dS(B_\parallel)$ for the 0-1 and 2-3 electron transitions.  The 2-electron ground state is a spin singlet, while the 3-electron ground state adds an additional unpaired electron, so both $dS_{0 \rightarrow 1}$ and $dS_{2 \rightarrow 3}$ correspond to transitions from total spin zero $(S_0=S_2=0)$ to total spin one-half, and one might expect identical field-dependent behaviors for $dS_{0\rightarrow 1}$ and $dS_{2\rightarrow 3}$ as long as $g \mu_{B} B_\parallel$ is less than the excited state energies.

The entropies of 1- and 3-electron states are suppressed when the Zeeman splitting lifts the spin degeneracy of the unpaired electron, following the Gibbs entropy for a two-level system:
%
\begin{align}
\label{eqn:gibbs}
        S &= k_B \sum_{i=\pm} p_{i}(B_\parallel, T) \ln{ p_{i}(B_\parallel,T) }
\end{align}
%
where $p_{\pm}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_{\parallel}}{k_B T}})^{-1}$ are the probabilities for the system to be in $+/-$ (up/down) spin states at a given field and temperature. Fits to the lineshape in Eqn.~\ref{eqn:gibbs} are shown in Fig.~\ref{fig:fig3}a for both transitions, allowing for peak height, a vertical offset, and the ratio $g/T$ as free parameters.

The entropy measurement agrees quantitatively with the expected $k_{B} \ln{2}$ at $B_\parallel=0$, collapsing to zero at high field when spin degeneracy is broken.  Using a $T=T_{res}$ of 130 \SI{100}{\milli\kelvin} that is midway between the low ($T_{MC}$) and high ($T_{MC}+dT$) points of the temperature oscillation yields $g=0.43$ or $g=0.4$ for the $0\rightarrow 1$ and $2\rightarrow 3$ transitions respectively, consistent with the value $g=-0.44$ expected for GaAs and measured via bias spectroscopy in a separate measurement (Figs.~\ref{fig:fig3}bc).  At a qualitative level, the high-field crossover to a perfectly antisymmetric lineshape for $dg_{sens}(V_G)$ at the charge transition clearly indicates $\delta$ from Eqn.~\ref{eqn:dg} going to zero, corresponding to no entropy change between the two charge states (Figs.~\ref{fig:fig3}de).

@@

The entropy of the 0 and 2-electron states are both zero at low magnetic field, so $dS$ for the $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}a) can be understood as the inverse of Fig \ref{fig:fig3}a as long as $B_\parallel$ is less than around \SI{5}{\tesla}.  There, the 2-electron ground state remains in the spin singlet state with zero entropy while 1-electron ground state entropy goes from $k_B\ln{2}$ to zero due to Zeeman splitting. At higher field, the 1-electron ground state remains non-degenerate while the 2-electron ground state becomes two-fold degenerate where the singlet ($S$) and triplet ($T_+$) states cross. The singlet-triplet crossing is seen clearly around at \SI{9}{\tesla} in bias spectroscopy data for the $1\rightarrow 2$ transition (Fig.~\ref{fig:fig4}bc).

The field-dependent entropy measurement for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities as before for the 1-electron states and $p_{S/T}(B_\parallel, T) = (1+ e^{\mp \frac{g\mu_B B_\parallel - \Delta_{ST}}{k_B T}})^{-1}$ for the 2-electron singlet-triplet states, where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. Fitting the data in Fig.~\ref{fig:fig4}c we find $g(B{=}0T)=-0.46$, $g(B{=}9T)=-0.31$, $\Delta_{ST}(B{=}9T)$ = \SI{240}{\micro\electronvolt}, $dS(B{=}0T) = 1.04\ln{2}$, and $dS(B{=}9T) = 1.01\ln{2}$. Each of these values is consistent with data in Figs.~\ref{fig:fig4}a and b as well as previous GaAs dot measurements [cite why g should change in high field]. This fit confirms that the $dS$ signal can be related directly to a tunable degeneracy in the system.

@@

\noindent \textbf{Methods} The device was built on a AlGaAs/GaAs heterostructure, hosting a 2D electron gas with density and mobility at \SI{300}{\milli\kelvin} (determined on a separate chip) of \SI{2.42e11}{\per\square\centi\metre} and \SI[per-mode=symbol]{2.56e6}{\square\centi\metre\per\volt\per\second}.   Mesas and NiAuGe ohmic contacts to the 2DEG were defined by standard photolithography techniques, then \SI{10}{\nano\metre} of $\mathrm{HfO_2}$ was deposited via atomic layer deposition to improve the gating stability to the device. Electron beam lithography followed by evaporation of \SI{20}{\nano\metre} of Ti/Au defined the fine gate structures, including the \SI{200}{\nano\metre} diameter quantum dots. The measurement was carried out in an dilution refrigerator with a base temperature of \SI{14}{\milli\kelvin}.
